---
title: "EDA"
output: pdf_document
date: "2026-02-25"
---

```{r}
# Load required libraries
library(tidyverse)
library(ggplot2)
# Read the data
data <- read.csv("Sports_performance_target.csv")

# Convert categorical variables to factors
data$home_game   <- as.factor(data$home_game)
data$position    <- as.factor(data$position)
data$game_timing <- as.factor(data$game_timing)
data$met_target  <- as.factor(data$met_target)
```

Check whether the class, which is the variable met_target, is imbalanced in this dataset. If so, data preprocessing is required, otherwise the model will learn the pattern from the majority more than the minority. 

EDA 1 — met_target imbalance check
```{r}
ggplot(data, aes(x = met_target)) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count",
            aes(label = paste0(round(..count../sum(..count..)*100, 1), "%")),
            vjust = -0.5) +
  labs(title = "Fig 1: Class Distribution of Met Target",
       x = "Met Target (0 = No, 1 = Yes)",
       y = "Count") +
  theme_minimal()
```

Check the distribution of the continous variable, mainly see whether the data is skewed and any further action is needed.

EDA 2 - distributon of continuous var
```{r}
continuous_vars <- data %>%
  select(minutes_played, opponent_strength, travel_hours)

continuous_long <- pivot_longer(continuous_vars,
                                cols = everything(),
                                names_to = "Variable",
                                values_to = "Value")

ggplot(continuous_long, aes(x = Value)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_wrap(~ Variable, scales = "free") +
  labs(title = "Fig 2: Distribution of Continuous Variables",
       x = "Value",
       y = "Frequency") +
  theme_minimal()
```

Check multicollinearity, whether the 2 variables are multicollineared.

EDA 3a — opponent_strength vs minutes_played
```{r}
ggplot(data, aes(x = opponent_strength, y = minutes_played)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Fig 3a: Opponent Strength vs Minutes Played",
       x = "Opponent Strength",
       y = "Minutes Played") +
  theme_minimal()

cor(data$opponent_strength, data$minutes_played)
```

EDA 3b — travel_hours vs home_game
```{r}
ggplot(data, aes(x = home_game, y = travel_hours)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Fig 3b: Travel Hours by Home Game",
       x = "Home Game",
       y = "Travel Hours") +
  theme_minimal()
```

EDA 3c — travel_hours vs rest_days
```{r}
ggplot(data, aes(x = as.factor(rest_days), y = travel_hours)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Fig 3c: Travel Hours by Rest Days",
       x = "Rest Days",
       y = "Travel Hours") +
  theme_minimal()
```

Bias check, whether different position have different standard as meeting the target.

EDA 4 - position vs met_target (bias check)
```{r}
ggplot(data, aes(x = position, fill = met_target)) +
  geom_bar(position = "fill") +
  labs(title = "Fig 4: Proportion of Met Target by Position",
       x = "Position",
       y = "Proportion") +
  theme_minimal()
```

EDA 5 — 3-Variable Scatter Plot
X = Minutes Played
Y = Travel Hours
Color = Met Target

```{r}
ggplot(data, aes(x = minutes_played,
                 y = travel_hours,
                 color = met_target)) +
  geom_point(alpha = 0.7) +
  labs(title = "Fig 5: Minutes Played vs Travel Hours (Colored by Met Target)",
       x = "Minutes Played",
       y = "Travel Hours") +
  theme_minimal()
```

